---
title: "Untitled"
author: "Nathan Young"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(readr)
```

# Title
Your document should have a title that briefly summarizes your data analysis

## Synopsis  
Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Data processing  
The data for this project is obtained from the National Weather Service

For data import, we have two immediate tasks: first to download the `stormdata.csv` file from the source. 

```{r import_data, cache = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv")
```

And secondly to import the data into R. 

```{r, cache = TRUE, cache.extra = file.mtime("stormdata.csv")}
data <- read_csv("stormdata.csv", show_col_types = FALSE)
```

When `read_csv()` is used on the url to download the data directly, an error was generated that "Reading from remote 'bz2' compressed files is not supported, download the files locally first" leading to the two step process above. 



## Results  
### Which types of events are most harmful with respect to public health?  
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### Which types of events have the greatest economic consequences?  
